#install.packages("UsingR")
library(UsingR)
library(ggplot2)
library(reshape2)
data(galton)
long <- melt(galton)
g <- ggplot(long, aes(x= value , fill= variable))
g <- g + geom_histogram(color = "black", binwidth=1)
g <- g + facet_grid(.~variable)
g


```{r , fig.height=4, fig.width=4, fig.align='center'}
g <- ggplot(galton, aes(x = child)) + geom_histogram(fill = "salmon", colour = "black", binwidth=1)
g <- g + geom_vline(xintercept = mean(galton$child), size = 3)
g
```

---
## Finding the middle via least squares
* Consider only the children's heights. 
  * How could one describe the "middle"?
  * One definition, let $Y_i$ be the height of child $i$ for $i = 1, \ldots, n = 928$, then define the middle as the value of $\mu$
  that minimizes $$\sum_{i=1}^n (Y_i - \mu)^2$$
* This is physical center of mass of the histrogram.
* You might have guessed that the answer $\mu = \bar Y$.

```{r}
# We can play around with the slider and as we move futher from the mean in any direction the MSE gets bigger
library(manipulate)
myHist <- function(mu){
    mse <- mean((galton$child - mu)^2)
    g <- ggplot(galton, aes(x = child)) + geom_histogram(fill = "salmon", colour = "black", binwidth=1)
    g <- g + geom_vline(xintercept = mu, size = 3)
    g <- g + ggtitle(paste("mu = ", mu, ", MSE = ", round(mse, 2), sep = ""))
    g
}
manipulate(myHist(mu), mu = slider(62, 74, step = 0.5))
```

---
## The least squares est. is the empirical mean
```{r , fig.height=4, fig.width=4, fig.align='center'}
g <- ggplot(galton, aes(x = child)) + geom_histogram(fill = "salmon", colour = "black", binwidth=1)
g <- g + geom_vline(xintercept = mean(galton$child), size = 3)
g
```
### The math (not required for the class) follows as:
$$ 
\begin{align} 
\sum_{i=1}^n \left(Y_i - \mu\right)^2 & = \
\sum_{i=1}^n \left(Y_i - \bar Y + \bar Y - \mu\right)^2 \\ 
& = \sum_{i=1}^n \left(Y_i - \bar Y\right)^2 + \
2 \sum_{i=1}^n \left(Y_i - \bar Y\right)  \left(\bar Y - \mu\right) +\
\sum_{i=1}^n \left(\bar Y - \mu\right)^2 \\
& = \sum_{i=1}^n \left(Y_i - \bar Y\right)^2 + \
2 \left(\bar Y - \mu\right) \sum_{i=1}^n \left(Y_i - \bar Y\right) +\
\sum_{i=1}^n \left(\bar Y - \mu\right)^2 \\
& = \sum_{i=1}^n \left(Y_i - \bar Y\right)^2 + \
2 \left(\bar Y - \mu\right)  \left(\left(\sum_{i=1}^n Y_i\right) -\
 n \bar Y\right) +\
\sum_{i=1}^n \left(\bar Y - \mu\right)^2 \\
& = \sum_{i=1}^n \left(Y_i - \bar Y\right)^2 + \
 \sum_{i=1}^n \left(\bar Y - \mu\right)^2\\ 
& \geq \sum_{i=1}^n \left(Y_i - \bar Y\right)^2 \
\end{align} 
$$

---
## Comparing childrens' heights and their parents' heights
## primary failing is that data is overplotted
```{r, dependson="galton",fig.height=4,fig.width=4, fig.align='center'}
ggplot(galton, aes(x = parent, y = child)) + geom_point()
```

---
better plot
Size of point represents number of points at that (X, Y) combination (See the Rmd file for the code).

```{r freqGalton, dependson="galton",fig.height=6,fig.width=7,echo=FALSE}
library(dplyr)
freqData <- as.data.frame(table(galton$child, galton$parent))
names(freqData) <- c("child", "parent", "freq")
freqData$child <- as.numeric(as.character(freqData$child))
freqData$parent <- as.numeric(as.character(freqData$parent))
g <- ggplot(filter(freqData, freq > 0), aes(x = parent, y = child))
g <- g  + scale_size(range = c(2, 20), guide = "none" )
g <- g + geom_point(colour="grey50", aes(size = freq+20, show_guide = FALSE))
g <- g + geom_point(aes(colour=freq, size = freq))
g <- g + scale_colour_gradient(low = "lightblue", high="white")                    
g
```

---
## Regression through the origin
* Suppose that $X_i$ are the parents' heights.
* Consider picking the slope $\beta$ that minimizes $$\sum_{i=1}^n (Y_i - X_i \beta)^2$$
* This is exactly using the origin as a pivot point picking the
line that minimizes the sum of the squared vertical distances
of the points to the line
* Use R studio's  manipulate function to experiment
* Subtract the means so that the origin is the mean of the parent
and children's heights

---
```{r, echo = TRUE, eval = FALSE}
y <- galton$child - mean(galton$child)
x <- galton$parent - mean(galton$parent)
freqData <- as.data.frame(table(x, y))
names(freqData) <- c("child", "parent", "freq")
freqData$child <- as.numeric(as.character(freqData$child))
freqData$parent <- as.numeric(as.character(freqData$parent))
myPlot <- function(beta){
    g <- ggplot(filter(freqData, freq > 0), aes(x = parent, y = child))
    g <- g  + scale_size(range = c(2, 20), guide = "none" )
    g <- g + geom_point(colour="grey50", aes(size = freq+20, show_guide = FALSE))
    g <- g + geom_point(aes(colour=freq, size = freq))
    g <- g + scale_colour_gradient(low = "lightblue", high="white")                     
    g <- g + geom_abline(intercept = 0, slope = beta, size = 3)
    mse <- mean( (y - beta * x) ^2 )
    g <- g + ggtitle(paste("beta = ", beta, "mse = ", round(mse, 3)))
    g
}
manipulate(myPlot(beta), beta = slider(0.6, 1.2, step = 0.02))
```

---
## The solution 
### In the next few lectures we'll talk about why this is the solution
```{r}
lm(I(child - mean(child))~ I(parent - mean(parent)) - 1, data = galton)
```

---
```{r, fig.height=6,fig.width=7,echo=FALSE}
freqData <- as.data.frame(table(galton$child, galton$parent))
names(freqData) <- c("child", "parent", "freq")
freqData$child <- as.numeric(as.character(freqData$child))
freqData$parent <- as.numeric(as.character(freqData$parent))
g <- ggplot(filter(freqData, freq > 0), aes(x = parent, y = child))
g <- g  + scale_size(range = c(2, 20), guide = "none" )
g <- g + geom_point(colour="grey50", aes(size = freq+20, show_guide = FALSE))
g <- g + geom_point(aes(colour=freq, size = freq))
g <- g + scale_colour_gradient(low = "lightblue", high="white")                    
lm1 <- lm(galton$child ~ galton$parent)
g <- g + geom_abline(intercept = coef(lm1)[1], slope = coef(lm1)[2], size = 3, colour = grey(.5))
g
```

```{r}
lm(I(child - mean(child))~ I(parent - mean(parent)) - 1, data = galton)
#Call:
lm(formula = I(child - mean(child)) ~ I(parent - mean(parent)) - 
    1, data = galton)

#Coefficients:
I(parent - mean(parent))  
                   0.646  
```